---
title: "Retail Store Sales Dataset Documentation"
output: pdf_document
date: "2025-12-02"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
```

# **1. Introduction**

This document provides a short documentation of the dataset used for the project.
The dataset was obtained from Kaggle:

**Source:** *Retail Store Sales (Dirty) – For Data Cleaning*
URL: [https://www.kaggle.com/datasets/ahmedmohamed2003/retail-store-sales-dirty-for-data-cleaning/data](https://www.kaggle.com/datasets/ahmedmohamed2003/retail-store-sales-dirty-for-data-cleaning/data)

The goal of this documentation is to summarize the dataset, describe its variables and types, and provide a reproducible R script used to load the data.

# **2. Dataset Description**

The dataset contains transactional retail sales data including product category, pricing, customer information, and purchase behavior.
It is intended for data cleaning and preprocessing tasks.

```{r load-data}
# Adjust filename if needed
data <- read_csv("retail_store_sales.csv")

# Show first few rows
head(data)
```

# **3. Summary of the Dataset**

## **3.1 Number of Rows and Columns**

```{r dims}
dim(data)
```

## **3.2 Summary of Variables**

```{r summary}
summary(data)
```

## **3.3 Data Types of Each Variable**

```{r structure}
glimpse(data)
```

# **4. Variable Description**

| Variable Name        | Description                                         |
| -------------------- | --------------------------------------------------- |
| **Transaction ID**   | Unique identifier for each transaction              |
| **Customer ID**      | Unique identifier for each customer                 |
| **Category**         | Product category (e.g., Patisserie, Butchers)       |
| **Item**             | Specific purchased item                             |
| **Price Per Unit**   | Price of a single item unit                         |
| **Quantity**         | Quantity purchased                                  |
| **Total Spent**      | Price × Quantity                                    |
| **Payment Method**   | Mode of payment (Digital Wallet, Credit Card, etc.) |
| **Location**         | Purchase location (Online or physical store)        |
| **Transaction Date** | Date of the transaction                             |
| **Discount Applied** | Boolean value indicating if a discount was used     |

# **5. Example Data (First Few Observations)**

```{r head}
head(data, 5)
```

# **6. Conclusion**

This dataset provides a structured but intentionally "dirty" retail transaction record intended for data cleaning exercises. The documentation summarizes its structure and provides reproducible R code for loading and exploring the data.

# **7. R Script for Loading the Data (for Lecturer)**

Save the following as **load_data.R** and upload along with your dataset:

```r
library(readr)

data <- read_csv("retail_store_sales.csv")

print(head(data))
print(str(data))
```


