---
title: "Student Performance Dataset Documentation"
output: pdf_document
mainfont: "Calibri"
date: "2025-12-27"
---
## Team
Melisa Cihan, Hrusheekesh Sawarkar, Kezia Fernandes, Raju Ahmed

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# required packages
library(readr)
library(dplyr)
```

# **1. Introduction**

This document provides a short documentation on the student performance dataset used for the project.
The dataset was obtained from Machine Learning Repository:

**Source:** *Retail Store Sales (Dirty) – For Data Cleaning*
URL: [https://archive.ics.uci.edu/dataset/320/student+performance](https://archive.ics.uci.edu/dataset/320/student+performance)

The goal of this documentation is to summarize the dataset, describe its variables and types and provide a reproducible R script used to load the data.

# **2. Dataset Description**

The dataset contains student performance data in secondary education of two Portugese schools.
Two datasets are provided regarding the performance in two distinct subjects: Mathematics (mat) and Portuguese language (por). However, we will only analyze and focus on the dataset for Mathematics (mat). Since this dataset contains too many variables we will only consider a subset of these variables, which are displayed in the following.

```{r load-data}
# required packages
library(readr)
library(dplyr)

data <- read_csv("/Users/hrusheekeshsawarkar/Projects/StatsProj/Student_Performance/student-mat.csv")


```

# **3. Summary of the Dataset**

## **3.1 Number of Rows and Columns**

```{r dims}
dim(data)
```

## **3.2 Summary of Variables**

```{r summary}
summary(data)
```

## **3.3 Data Types of Each Variable**

```{r structure}
glimpse(data)
```

# **4. Variable Description**

| Variable Name | Data Type | Scale of Measure | Description |
|---------------|-----------|------------------|-------------|
| **sex** | Binary | Nominal | Student's sex (F - female, M - male) |
| **age** | Numeric | Numeric discrete | Student's age (ranges from 15 to 22 years) |
| **Medu** | Categorical | Ordinal | Mother's education level (0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education, 4 - higher education) |
| **traveltime** | Categorical | Ordinal | Home to school travel time (1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, 4 - >1 hour) |
| **studytime** | Categorical | Ordinal | Weekly study time (1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, 4 - >10 hours) |
| **failures** | Numeric | Discrete | Number of past class failures (numeric value, where 4 represents 3 or more failures) |
| **paid** | Binary | Nominal | Extra paid classes within the course subject (yes or no) |
| **activities** | Binary | Nominal | Participation in extra-curricular activities (yes or no) |
| **higher** | Binary | Nominal | Wants to pursue higher education (yes or no) |
| **internet** | Binary | Nominal | Internet access at home (yes or no) |
| **famrel** | Categorical | Ordinal | Quality of family relationships (scale from 1 - very bad to 5 - excellent) |
| **absences** | Numeric | Discrete | Number of school absences (ranges from 0 to 93) |
| **G3** | Numeric | Continuous* | Final grade (ranges from 0 to 20) |

# **5. Example Data (First Few Observations)**

```{r head}
head(data, 5)
```

# **6. Conclusion**

# **7. Limitation**
This dataset does not contain any continuous variables. Even variables like traveltime are displayed as categorical variables, since this is generally the case in social sciences. However, for some of the analyses we would like to use the variable G3 (final grade, ranges from 0-20) as it were a continuous variable, even though it displays discerete integer values. 
